\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-01-01 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf AE.2. Comments by AE}:\\
{\em Authors are also encouraged to revisit their overview of prior art.}

{\nib {\blue As suggested in the comments, the revised manuscript has expanded the overview of prior art. Specifically, we have included \ul{two new references [22], [23] on stochastic differential algebraic equations (SDAEs) and cited them inline in Sec. I.1, second paragraph, as detailed in our response to the comment \textbf{R3.2} in this document}. This includes the reference suggested in comment \textbf{R1.3}. \ul{We have also included the reference suggested in comment \textbf{R4.4}; this is reference [28] in the revised manuscript. Its inline citation is in Sec. I.1, last paragraph, first sentence}.}}

\noindent
{\bf AE.3. Comments by AE}:\\
{\em Presentation of technical points and results can also be improved.}

{\nib {\blue \ul{Following the comment \textbf{R4.3} to improve the presentation, in the revised manuscript, we have expanded the last two paragraphs of Sec. III-A with a new reference [50]}. Please see our response to that comment below for details.} {\red{TBD on figure clarity}}
}

@@

\noindent
{\bf R3.3. Comments by reviewer}:\\
{\em There are several typos in the test (especially in the references).}

{\nib \blue{TBD. We corrected the two typos pointed out in comment \textbf{R4.8} in this rebuttal document.}}

\noindent
{\bf R3.4. Comments by reviewer}:\\
{\em Figures 5 and 6 are unreadable.}

{\nib \blue{In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, Figures 5 and 6 are made larger. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6).}}

### 2022-01-01 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf R3.3. Comments by reviewer}:\\
{\em There are several typos in the test (especially in the references).}

{\nib \blue{TBD. We corrected the two typos pointed out in comment \textbf{R4.8} in this rebuttal document.}}

\noindent
{\bf R3.4. Comments by reviewer}:\\
{\em Figures 5 and 6 are unreadable.}

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$, thanks to the periodic boundary condition on the angular variable. Our philosophy of technical communication is that the onus is on the authors, not on the readers, to present the scientific information as correctly as possible. And yes, we are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to understand the statistical meaning conveyed in those plots. We prefer ours  not be another paper contributing to the accrual of such.}}

\begin{figure}[t]
\centering
\includegraphics[width=0.99\linewidth]{IEEE14marginals.png}
\caption{\small{Example $(\theta_i,\omega_i)$ bivariate marginals, for the IEEE 14 bus system's generators $i=1,\hdots, 5$, shown as planar contour plots. The columns correspond to time snapshots; the rows correspond to the five generators. The colorbar corresponds to the value of the bivariate marginals. The simulation set up is described in Sec. V-A of the manuscript.}}
\vspace*{-0.1in}
\label{fig:PlanarContourBivariateMarginals}
\end{figure}

### 2022-01-01 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\nib \blue{Thank you for the kind and encouraging remarks.}}

\noindent
{\bf R2.2. Comments by reviewer}:\\
{\em The paper's language and presentation make it seem like it is more oriented towards the applied mathematics and control communities. It is highly suggested that the authors revise the exposition of ideas and concepts, putting more emphasis on the power system so that it can be well-understood by the power system experts.}

{\nib \blue{We appreciate the comment. \ul{In the revised manuscript, we have made some changes toward this direction. This includes slightly expanded overview of the related power system literature (see the two sentences added in Sec. I.1, second paragraph with two new references [22], [23]), citing recent reference [28] in Sec. I.1, third paragraph. Please note that the literature review in Sec. I.1 is completely based on the power system literature}.

Having said that, we believe our technical contribution is fundamentally interdisciplinary in the sense it is not mere translational research. In other words, we are not simply importing ideas from a different domain and applying them to a power system problem. Instead, we make novel contributions at the (possibly unanticipated) intersection of two areas: measure-valued gradient flows and power system dynamics. The presentation of the paper reflects the same, perhaps rightfully so. Notice in particular that our presentation is bottom-up: we start from power system literature review and mention that the problem addressed in our paper was already recognized as a technical challenge in that literature. We move onto the theoretical ideas and tools much later in the manuscript, as the build up demands.

@@

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$, thanks to the periodic boundary condition on the angular variable. Our philosophy of technical communication is that the onus is on the authors, not on the readers, to present the scientific information as correctly as possible. And yes, we are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to understand the statistical meaning conveyed in those plots. We prefer ours  not be contributing to the accrual of such pain, and instead make plots that depict the variables as they are.}}

### 2022-01-01 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf AE.2. Comments by AE}:\\
{\em Authors are also encouraged to revisit their overview of prior art.}

{\nib {\blue As suggested in the comments, the revised manuscript has expanded the overview of prior art. Specifically, we have included \ul{two new references [22], [23] on stochastic differential algebraic equations (SDAEs) and cited them inline in Sec. I.1, second paragraph, as detailed in our response to the comment \textbf{R3.2} in this document}. This includes the reference suggested in comment \textbf{R1.3}. \ul{We have also included the reference suggested in comment \textbf{R4.4}; this is reference [28] in the revised manuscript. Its inline citation is in Sec. I.1, last paragraph, first sentence}.}}

\noindent
{\bf AE.3. Comments by AE}:\\
{\em Presentation of technical points and results can also be improved.}

{\nib {\blue \ul{Following the comment \textbf{R4.3} to improve the presentation, in the revised manuscript, we have expanded the last two paragraphs of Sec. III-A with a new reference [50]}. Please see our response to that comment below for details.

\ul{As detailed in the response to comment \textbf{R3.4}, in the revised manuscript, we have enlarged Figures 5 and 6, and expanded the plot in Fig. 6}. There we also explained why those plot styles are used.

\ul{Furthermore, Fig. 7 in the revised manuscript has also been expanded to include the univariate angular velocity transient marginal plots for all the five generators of the IEEE 14 bus system; previous version had this plot only for the generator 4 (buss 6)}.

{\red{New plots}}
}}

@@

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$, thanks to the periodic boundary condition on the angular variable. Our philosophy of technical communication is that the onus is on the authors, not on the readers, to clarify the scientific information as correctly as possible. And yes, we are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to understand the statistical meaning conveyed in those plots. We prefer ours  not be contributing to the accrual of such pain, and instead make plots that depict the variables as they are.}}

### 2022-01-02 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{\large \bf Response to Reviewer \#3:}

\noindent
{\bf R3.1. Comments by reviewer}:\\
{\em My main concern with this work is that all its derivations are deeply model-dependent and, unfortunately, the model that the author consider is not adequate to represent modern power systems. Moreover, second order model of synchronous machines are adequate only in a time scale of few seconds, so, even if the system were effectively including only conventional power plants, the model would be acceptable only for very short-term transients, i.e., transients for which the propagation of the noise is immaterial. In other words, the proposed model is useless.}

{\nib \blue{The reviewer's point is well taken in that the second order model may not be adequate to capture all the complexities of modern power systems, and }}

@@

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$ rad, thanks to the periodic boundary condition on the angular variable. Our philosophy of technical communication is that the onus is on the authors, not on the readers, to clarify the scientific information as correctly as possible. And yes, we are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to decipher the statistical meaning conveyed in those plots. We prefer ours  not be contributing to the accrual of such pain, and instead make plots that depict the variables as they are.}}

@@

{\nib \blue{TBD.}}

\noindent
{\bf R4.3. Comments by reviewer}:\\
{\em Regarding the presentation, the reviewer suggests making the paper more self-contained. Some discussion in this paper highly relies on the authors' previous work, which can be improved by either adding a summarization or trimming out the materials.

Following the same vein, there are some unnecessary information and technical points provided in this paper that may be distracting.
For example, on page 3, why would the authors mention the Kuramoto oscillator model?
On page 4, is the detailed discussion regarding Wasserstein-2 metric necessary? Similarly, on page 5, the discussion about the natural energy functional and how the gradient drift case does not apply also seems too detailed.}

{\nib \blue{We have followed the reviewer's suggestion to improve the presentation. \ul{Specifically, in the revised manuscript's Sec. III-A, last two paragraphs summarize the technical relation of this paper w.r.t. the authors' previous work. In particular, we have added three more sentences with a new reference. To improve readability, have also broken what was a single paragraph in the previous version, into two separate paragraphs.}

@@

We appreciate the reviewer's point that the sentence containing equation (6) and the mention of Kuramoto model can be removed without affecting the rest of the development. However, we prefer to keep it since it adds didactic value, and many engineers may feel comfortable by the explicit mention that (1) is simply a rewriting of (6). Furthermore, an astute reader, encountering that sentence will realize that the scope of the technical contribution of this paper is broad because the algorithmic framework developed here will be applicable to several other non-power system applications (mentioned in the two cited references in that sentence) where the second order Kuramoto model is used. Since this is just a matter of one sentence, we will prefer to keep it. We hope you approve.

\ul{On the other hand, the detailed discussion regarding Wasserstein gradient flow and why that standard development cannot be applied in this case, is in fact, the heart of this paper}. Please note that \ul{ours is not a recipe paper} where a numerical algorithm is proposed backed by some theory and simulations. \ul{The key contribution of this paper, from the authors' viewpoint, is to help distill out the core technical difficulty that hinders simply importing standard techniques from the optimal transport literature}. We believe this clarification of what really is the technical crux of the matter--versus what are peripheral issues--should be illuminating to not just power systems researchers, but also to applied mathematicians underscoring how the standard theory and algorithms for measure-valued gradient flows need to be adapted in engineering applications.}}

### 2022-01-02 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{\large \bf Response to Reviewer \#3:}

\noindent
{\bf R3.1. Comments by reviewer}:\\
{\em My main concern with this work is that all its derivations are deeply model-dependent and, unfortunately, the model that the author consider is not adequate to represent modern power systems. Moreover, second order model of synchronous machines are adequate only in a time scale of few seconds, so, even if the system were effectively including only conventional power plants, the model would be acceptable only for very short-term transients, i.e., transients for which the propagation of the noise is immaterial. In other words, the proposed model is useless.}

{\nib \blue{The reviewer's point is well taken in the sense the second order model may not be adequate to capture all the complexities of modern power systems, and further research is needed to design/extend nonparametric joint PDF propagation methods such as the one we propose with higher order generator dynamics. \ul{Please note that the last sentence of our Conclusion section explicitly stated so}.

At the same time, it is important to understand how exactly this paper is making a significant stride w.r.t. the current state-of-the-art even with the .}}

### 2022-01-02 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf AE.2. Comments by AE}:\\
{\em Authors are also encouraged to revisit their overview of prior art.}

{\nib {\blue As suggested in the comments, the revised manuscript has expanded the overview of prior art. Specifically, we have included \ul{two new references [22], [23] on stochastic differential algebraic equations (SDAEs) and cited them inline in Sec. I.1, second paragraph, as detailed in our response to the comment \textbf{R3.2} in this document}. This includes the reference suggested in comment \textbf{R1.3}. \ul{We have also included the reference suggested in comment \textbf{R4.4}; this is reference [28] in the revised manuscript. Its inline citation is in Sec. I.1, last paragraph, first sentence}.}}

\noindent
{\bf AE.3. Comments by AE}:\\
{\em Presentation of technical points and results can also be improved.}

{\nib {\blue We have made several updates to clarify the technical issues and to improve the presentation.

\ul{Following the comment \textbf{R4.3} to improve the presentation, in the revised manuscript, we have expanded the last two paragraphs of Sec. III-A with a new reference [50]}. Please see our response to that comment below for details.

@@

\newpage
\section*{\large \bf Response to Reviewer \#3:}

\noindent
{\bf R3.1. Comments by reviewer}:\\
{\em My main concern with this work is that all its derivations are deeply model-dependent and, unfortunately, the model that the author consider is not adequate to represent modern power systems. Moreover, second order model of synchronous machines are adequate only in a time scale of few seconds, so, even if the system were effectively including only conventional power plants, the model would be acceptable only for very short-term transients, i.e., transients for which the propagation of the noise is immaterial. In other words, the proposed model is useless.}

{\nib \blue{The reviewer's point is well taken in the sense the second order model may not be adequate to capture all the complexities of modern power systems, and further research is needed to design/extend nonparametric joint PDF propagation methods such as the one we propose with higher order generator dynamics. \ul{Please note that the last sentence of our Conclusion section explicitly stated so}.

\ul{At the same time, it is important to understand how exactly this paper is making a significant stride w.r.t. the current state-of-the-art even with the second order dynamics}. The last paragraph of our Sec. II-B clearly asserts that the state-of-the-art in directly computing the joint PDF remains the ref. [45] published in this journal in 2013, which considered the second order dynamics for the single machine infinite bus case via finite element discretization. In the same paragraph, we also cited two 2018 references [16, 17] published again in this journal, which did not directly propagate the state PDFs and instead computed some proxy for the same. \ul{So the key  question becomes: if the power system literature has already acknowledged the importance of solving this problem, why is there no progress since [45] for even moderate $n$ and even assuming seemingly simple second order models? What exactly is the fundamental impediment standing in the way of nonparametric computation of the joint state PDF subject to the power system dynamics}? Our Sec. I.2 clearly lists the challenges for the same. The crux is that addressing this challenge is not a matter of innovating on a better discretization scheme or innovating on computing architecture. \ul{See specifically the last two sentences of our Sec. II-B}. \ul{This paper is breaking the stalemate by fundamentally and rigorously innovating on what it really means to solve the underlying PDE, albeit subject to the second order dynamics}. Of course, we expect follow up research to extend the proposed approach for higher order models but we cannot have progress with a complex model unless we first demonstrate how to tackle the simpler second order case -- a case that the existing literature cannot tackle for nonparametric online PDF computation.

\ul{Last but not the least, we find the reviewer's assertion ``In other words, the proposed model is useless" vacuous for the following reasons}.

(i) We \ul{do not propose} the second order model. We use the second order SDE that albeit simple, is widely used in power system dynamic simulation, as the sample path dynamics. In this paper, we are not interested in simulating sample paths. Our interest is in nonparametrically computing the transient joint PDFs subject to the given sample path dynamics in the form of It\^{o} SDE. Our usage of this particular sample path dynamics does not preempt the possibility of designing a variant of the proposed proximal recursion for performing computation of the same nature when the given It\^{o} SDE is of higher order. Indeed, this is what we alluded to in the last sentence of the Conclusions section.

(ii) \ul{What we propose is not a model}. Transcribing the Fokker-Planck operator into a generalized gradient flow, and thereby deriving the corresponding proximal recursion, is a purely mathematical construct. There are no modeling assumptions made in that derivation. Our contribution in this paper is to initiate and showcase this new way of solving the transient Fokker-Planck PDE subject to a simple but popular power system dynamic model. Extending this framework to higher order sample path dynamics is mentioned as future work, and is beyond the scope of this particular paper.

### 2022-01-03 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\newpage
\section*{\large \bf Response to Reviewer \#3:}

\noindent
{\bf R3.1. Comments by reviewer}:\\
{\em My main concern with this work is that all its derivations are deeply model-dependent and, unfortunately, the model that the author consider is not adequate to represent modern power systems. Moreover, second order model of synchronous machines are adequate only in a time scale of few seconds, so, even if the system were effectively including only conventional power plants, the model would be acceptable only for very short-term transients, i.e., transients for which the propagation of the noise is immaterial. In other words, the proposed model is useless.}

{\nib \blue{The reviewer's point is well taken in the sense the second order model may not be adequate to capture all the complexities of modern power systems, and further research is needed to design/extend nonparametric joint PDF propagation methods such as the one we propose with higher order generator dynamics. \ul{Please note that the last sentence of our Conclusion section explicitly stated so}.

\ul{At the same time, it is important to understand how exactly this paper is making a significant stride w.r.t. the current state-of-the-art even with the second order dynamics}. The last paragraph of our Sec. II-B clearly asserts that the state-of-the-art in directly computing the joint PDF remains the ref. [45] published in this journal in 2013, which considered the second order dynamics for the single machine infinite bus case via finite element discretization. In the same paragraph, we also cited two 2018 references [16, 17] published again in this journal, which did not directly propagate the state PDFs and instead computed some proxy for the same. \ul{So the key  question becomes: if the power system literature has already acknowledged the importance of solving this problem, why is there no progress since [45] for even moderate $n$ and even assuming seemingly simple second order models? What exactly is the fundamental impediment standing in the way of nonparametric computation of the joint state PDF subject to the power system dynamics}? Our Sec. I.2 clearly lists the challenges for the same. The crux is that addressing this challenge is not a matter of innovating on a better discretization scheme or innovating on computing architecture. \ul{See specifically the last two sentences of our Sec. II-B}. \ul{The present paper is breaking the stalemate by fundamentally and rigorously innovating on what it really means to solve the underlying PDE, albeit subject to the second order dynamics}. As we have stated up front, we expect follow up research to extend the proposed approach for higher order models but we cannot have progress with a complex model unless we first demonstrate how to tackle the simpler second order case -- a case that the existing literature cannot tackle for nonparametric online PDF computation.

@@

\ul{Last but not the least, we find the reviewer's assertion ``In other words, the proposed model is useless" vacuous for the following reasons}.

(i) We \ul{do not propose} the second order model. We use the second order SDE that albeit simple, is widely used in power system dynamic simulation, as the sample path dynamics. In this paper, we are not interested in simulating sample paths. Our interest is in nonparametrically computing the transient joint PDFs subject to the given sample path dynamics in the form of It\^{o} SDE. \ul{Our usage of this particular sample path dynamics does not preempt the possibility of designing a variant of the proposed proximal recursion for performing computation of the same nature when the given It\^{o} SDE is of higher order. Indeed, this is what we alluded to in the last sentence of the Conclusions section}.

(ii) \ul{What we propose is not a model}. Transcribing the Fokker-Planck operator into a generalized gradient flow, and thereby deriving the corresponding proximal recursion, is a purely mathematical construct. There are no modeling assumptions made in that derivation. \ul{Our contribution in this paper is to initiate and showcase this new way of solving the transient Fokker-Planck PDE subject to a simple but popular power system dynamic model}. Extending this framework to higher order sample path dynamics is mentioned as future work, and is beyond the scope of this particular paper.

### 2022-01-07 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

%\newpage
\spacingset{1}

\section*{\large \bf Response to AE:}

\noindent
{\bf AE.1. Comments by AE}:\\
{\em The authors are requested to revise and resubmit their paper after addressing reviewer feedback. Of particular importance is that reviewers point out the model leveraged for analysis is overly simplistic.}

{\nib {\blue We are indebted to the AE and the reviewers for their help to improve the manuscript.

In addition to the itemized responses to the reviewers' comments and the corresponding revisions marked in blue in the manuscript, we have made the \ul{following additional fixes and improvements} (also marked in blue in the revised manuscript):\\
(i) In equation (8), we corrected a typo in its right hand side: the second summand runs over pairs $(i,j)$ such that $i<j$.\\
(ii)

{\red{Regarding the usage of simple model, please see our response to comment \textbf{R3.1} and ?? In summary, the current state-of-the-art in power system literature for this problem is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\(i) The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for higher order synchronous machine models. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. The authors have been completely up front by explicitly mentioning this in the Conclusions section. We have also reiterated the same in this rebuttal.\\(ii) Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models. In fact, a significant contribution of our article, in the authors' opinion, is that it distills out what exactly is technically challenging and what are the peripheral issues.}} 
}}

@@

\newpage
\spacingset{1}

\section*{\large \bf Response to Reviewer \#1:}

\noindent
{\bf R1.1. Comment by reviewer}:\\
{\em The original contribution of the article is not clear.  Indeed, what is original new from this article, not the summary from existing techniques?}

{\nib {\blue The original contribution of this article is to design an algorithm for \emph{online nonparametric} computation of the joint state PDFs subject to stochastic uncertainties (initial conditions, parameters, process noise and combinations thereof) in power system dynamics.

The qualifier ``online" means that it is not a variant of Monte Carlo where samples are first propagated \ul{and then} a function approximation is performed offline to estimate the joint state PDF at a snapshot. The qualifier ``nonparametric" means that neither the statistics nor the dynamics are approximated \emph{a priori}. This is novel because to the best of the author's knowledge, there does not exist a power systems paper which can accomplish the same. The state-of-the-art in power system literature is discussed in Sec I.1. More importantly, Sec. I.2 details why the state-of-the-art is what it is. \ul{We specifically explain the contribution of this article in Sec. I.3 titled ``Contributions of this paper".}
}}

@@

\noindent
{\bf R1.2. Comments by reviewer}:\\
{\em The structure of the article should be highly improved. A large portion of the contexts is simply repeating the well-known power system classical model, kroon reduction, and the FPF equations.  Only from section IV, page 6, I can see some of the proposed weighted points methods.}

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea; the idea itself is new and consequently, needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is mistaken}. Notice that we clearly laid out the navigational chart of this paper in the paragraph right before Sec. II starts.}}

@@

\noindent
{\bf R2.4. Comments by reviewer}:\\
{\em Incorporation of power system dynamics into the proposed uncertainty propagation framework is a bit vague. The proposed study seems not to have utilized any tool capable of simulating power system dynamics. Could the authors explain the part pertaining to the dynamic simulation in more detail?}

{\nib \blue{In this paper, we are interested in propagating the joint PDF values of the states of a power system dynamic model subject to joint stochastic uncertainties. Our focus is on explicitly computing these transient joint state PDFs at the time-varying state vector samples, which amounts to co-evolving the states and the joint PDF values over time. \ul{This is different from Monte Carlo or the traditional density estimation techniques since they approximate the joint PDFs via post-processing, not as an online computation}. To the best of our knowledge, there does not exist a power system simulation tool which can do this online computation while making neither statistical nor dynamical approximation.

Please note that in Sec. V, we have used MATPOWER and the Python-library ANDES (with inline citation of refs. [62] and [63] therein) for obtaining some simulation parameters, as explained in the first paragraph of V-A.}}

@@

\ul{At the same time, it is important to understand how exactly this paper is making a significant stride w.r.t. the current state-of-the-art even with the second order dynamics}. The last paragraph of our Sec. II-B clearly asserts that the state-of-the-art in directly computing the joint PDF remains the ref. [45] published in this journal in 2013, which considered the second order dynamics for the single machine infinite bus case via finite element discretization. In the same paragraph, we also cited two 2018 references [16, 17] published again in this journal, which did not directly propagate the state PDFs and instead computed some proxy for the same. \ul{So the key  question becomes: if the power system literature has already acknowledged the importance of solving this problem, why is there no progress since [45] for even moderate $n$ and even assuming seemingly simple second order models? What exactly is the fundamental impediment standing in the way of nonparametric computation of the joint state PDF subject to the power system dynamics}? Our Sec. I.2 clearly lists the challenges for the same. The crux is that addressing this challenge is not a matter of innovating on a better discretization scheme or innovating on computing architecture. \ul{See specifically the last two sentences of our Sec. II-B}. \ul{The present paper is breaking the stalemate by fundamentally and rigorously innovating on what it really means to solve the underlying PDE, albeit subject to the second order dynamics}. As we have stated up front, we expect follow up research to extend the proposed approach for higher order models but we cannot have progress with a complex model unless we first demonstrate how to tackle the simpler second order case -- a case that the existing literature cannot tackle for nonparametric online PDF computation.

\ul{Last but not the least, we find the reviewer's assertion ``In other words, the proposed model is useless" moot for the following reasons}.

### 2022-01-07 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\nib {\blue The original contribution of this article is to design an algorithm for \emph{online nonparametric} computation of the joint state PDFs subject to stochastic uncertainties (initial conditions, parameters, process noise and combinations thereof) in power system dynamics.

The qualifier ``online" means that it is not a variant of Monte Carlo where samples are first propagated \ul{and then} a function approximation is performed offline to estimate the joint state PDF at a snapshot. The qualifier ``nonparametric" means that neither the statistics nor the dynamics are approximated \emph{a priori}. This is novel because to the best of the author's knowledge, there does not exist a power systems paper which can accomplish the same. The state-of-the-art in power system literature is discussed in Sec I.1. More importantly, Sec. I.2 details why the state-of-the-art is what it is. \ul{We specifically explain the contribution of this article in Sec. I.3 titled ``Contributions of this paper".}
}}

\noindent
{\bf R1.2. Comments by reviewer}:\\
{\em The structure of the article should be highly improved. A large portion of the contexts is simply repeating the well-known power system classical model, kroon reduction, and the FPF equations.  Only from section IV, page 6, I can see some of the proposed weighted points methods.}

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea; the idea itself is new and consequently needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is mistaken}. Notice that we clearly laid out the navigational chart of this paper in the paragraph right before Sec. II starts.}}

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS_Rev1.tex
TPS  rebuttal round 1

Fig. \ref{fig:CompTimeIEEE14bus} shows the computational times needed to update $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ via the algorithm in Sec. \ref{sec:ProxAlgo}, for both \textbf{Cases I} and \textbf{II}.

>>>>

Fig. \ref{fig:CompTimeIEEE14bus} shows the computational times needed to update $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ via the algorithm in Sec. \ref{sec:ProxAlgo}, for both \textbf{Cases I} and \textbf{II}. {\blue{While the depiction of the computational times in Fig. \ref{fig:CompTimeIEEE14bus} appears unconventional, we next explain why that is meaningful in our framework.

At each fixed physical time $t_{k}=kh$, $k\in\mathbb{N}$, where $h$ is the fixed step-size, we perform the proximal update $\varrho_{k-1}\mapsto\varrho_{k}$ via Algorithm 1. This proximal update has no analytical solution and instead requires execution of a contractive fixed point recursion. The physical time $t_{k}$ is ``frozen" (zero-order-hold) during the fixed point recursions underlying this one-step proximal update. To claim that the proposed method is practical, we therefore, need to demonstrate that the computational timescale is faster than the physical timescale, i.e., computational times needed to complete the zero-order-hold fixed point recursions are orders-of-magnitude faster than $h$, the step-size of physical time. On the other hand, theoretical consistency (see the last paragraph of Sec. \ref{subsec:ProxUpdate}) requires that $h$ itself should be small. Thus, even if one has a provably convergent algorithm that can perform the one step proximal update, it is not clear \emph{a priori} if the computational time needed to do so can be much faster than $h$. If not, then the method would be statistically consistent but practically not useful. The Fig \ref{fig:CompTimeIEEE14bus} shows that the computational time needed to perform the proximal updates, for all $t_{k}$, are indeed orders-of-magnitude faster than $h$.
}}
